---
title: "Unsupervised learning - clustering and dimension reduction"
author: "Brian Chang"
date: "Fall 2019"
output:
  html_document:
    df_print: paged
fontsize: 11pt
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, eval = T)
```

# Lab Section
Download auto data from the *Statistical Learning* book website here: http://www-bcf.usc.edu/~gareth/ISL/data.html

Today, we are going over Hierarchical clustering, K-Means Clustering, PCA, and ICA. 

```{r load, include=FALSE}
library(ggplot2)
library(tidyverse)
library(ggfortify)
library(fastICA)
library(caret)
library(factoextra)
library(cluster)
library(FactoMineR)
```

```{r}
# read in Auto data
Auto_data <- read_csv("/Users/Brian Chang/SkyDrive/NYU/Fall 2019/Machine Learning/Homework/ML-hw1/Auto.csv")

#remove cars with unknown horsepower and set horsepower to numeric
Auto_data <- Auto_data %>% 
  filter(horsepower != "?") %>% 
  mutate(horsepower = as.numeric(horsepower)) %>% 
  as.data.frame()

#save car names 
Auto_data_names <- Auto_data$name

#data to cluster
Auto_data_clust <- Auto_data[,1:8]
dim(Auto_data_clust)

#392 is too much for a demo, so lets take the first 25
Auto_data_clust <- Auto_data_clust[1:25,]
rownames(Auto_data_clust) <- Auto_data_names[1:25]
```


## Hierarchical agglomerative clustering

Step 1. Assign each item to it's own cluster. We start with 25 clusters, one for each car. 

Step 2. Calculate a proximity matrix between each cluster.

Step 3. Find the pair of clusters closest to each other. 

Step 4. Merge these clusters and then recalculate similarity between clusters. Some options are: single linkage (distance is calculated from the nearest neighbors), complete linkage (distance is calculated from furthest neighbor), average linkage (distance is calculated from mean of different clusters). 

Step 5. Repeat Step 3 and 4 until there is only one cluster.

### In practice

Step 1. Each car is a cluster. 

Step 2. Create a distance matrix from Auto_data_clust.

```{r}
help("dist")
hierarchical_dist <- as.matrix(dist(Auto_data_clust, method = "euclidean"))
#View(hierarchical_dist)
```

Step 3. Find the two cars that are the most similar to each other and print the names of those two cars

```{r }
diag(hierarchical_dist) <- NA
arrayInd(which.min(hierarchical_dist), dim(hierarchical_dist))

#postitions 23 and 15 are the most similar. Lets go back to the names of the cars
Auto_data_names[23]
Auto_data_names[15]

```

Step 4. Merge the two clusters together using average linkage. 

```{r }
#replace pos 15 with the average of pos 15 and 23
hierarchical_dist[,15] <- apply((hierarchical_dist[,c(23,15)]),1,mean)
hierarchical_dist[15,] <- apply((hierarchical_dist[c(23,15),]),2,mean)

#remove pos 23
hierarchical_dist <- hierarchical_dist[-23,-23]

#now position 15 represents the cluster containing the saab99e and the toyota corona mark ii
```

Step 5. To complete the algorithm, go back to step 3 and iterate through all of the previous steps until there are no more rows left

```{r }
diag(hierarchical_dist) <- NA
arrayInd(which.min(hierarchical_dist), dim(hierarchical_dist))

#postitions 4 and 3 are the most similar
Auto_data_names[4]
Auto_data_names[3]

```

### R function 

Now that we know how the algorithm works, let's use the R function hclust. Plot the Dendogram resulting from clustering the Auto_data_clust using average linkage.  

```{r}
hierarchical_dist <- dist(Auto_data_clust, method = "euclidean")
tree <- hclust(hierarchical_dist, method="average")
plot(tree)
```

There is one more element to hierarchical clustering: Cutting the tree. Here, we can control how many clusters we want or the height of the tree. 
```{r}
#help(cutree)

# cut tree into 3 clusters
tree <- hclust(hierarchical_dist, method="average")
plot(tree)
tree_k2 <- cutree(tree, k = 2)
# plot the tree before running this line 
rect.hclust(tree, k = 3, h = NULL)
```


\newpage

## Principal Components Analysis (PCA)

Principal Components Analysis is a linear dimensionality reduction algorithm. If you want to learn more about linear algebra, I suggest the MIT Open Courseware class here : https://ocw.mit.edu/courses/mathematics/18-06-linear-algebra-spring-2010/
There are two ways of doing PCA, Single Value Decomposition (SVD), and the method we will use today, using the covariance matrix of the data. 

Step 1. Center data by subtracting the mean.

Step 2. Calculate covariance matrix of data.

Step 3. Perform Eigendecomposition of the covariance matrix. i.e. represent the matrix in terms of it's eigenvalues and eigen vectors

Step 4. Multiply the eigen vectors by the original data to express the data in terms of the eigen vectors. 

Step 1. Center the data by subtracting the mean of the each column from the values in that column

```{r}
Auto_data_clust_pca <- data.matrix(Auto_data_clust)

Center_auto <- apply(Auto_data_clust_pca, 2, function(x) x - mean(x))
```


Step 2. Calculate covariance matrix of the Auto data

```{r}

Covariance_auto <- cov(Center_auto)
```

Step 3.  Calculate eigen values and vectors

```{r}
Eigen_value_auto <- eigen(Covariance_auto)$value

#columns are the eigen vectors
Eigen_vector_auto <- eigen(Covariance_auto)$vector
```

Step 4. Multiply the eigen vector matrix by the original data. 

```{r}
PC <- as.data.frame(data.matrix(Center_auto) %*% Eigen_vector_auto)

ggplot(PC, aes(PC[,1], PC[,2])) + geom_point(aes(PC[,1], PC[,2]))
#+ geom_text(aes(label=Auto_data_names[1:8]), nudge_x = -2.5, nudge_y = 400)
```

Step 5. Find out which principal components explain the variance in the data. 

```{r}
#for each component, take the cumulative sum of eigen values up to that point and and divide by the total sum of eigen values
round(cumsum(Eigen_value_auto)/sum(Eigen_value_auto) * 100, digits = 2)
```

Principal component 1 and 2 explain 99.99 percent of the variance. Principal component 1,2, and 3 together explain 100% of the variance in the data. 

### R function 
Now that we know how PCA works, lets use the R funtion prcomp.

```{r}
help("prcomp")
autoplot(prcomp(Auto_data_clust_pca))
```

\newpage

## Independent Component Analysis (ICA)
ICA is an algorithm that finds components that are independent, subcomponents of the data. 

Step 1. Whiten the data by projecting the data onto the eigen vectors (PCA).

Step 2. Solve the X=AS equation by maximizing non-gaussianty in the variables(components) in S. 

This results in a matrix S with components that are independent from each other. 

We will use the fastICA algorithm.

First we will go backwards. 
Create a matrix S with the independent components
```{r}
#create two signals
S <- cbind(cos((1:500)/10), ((500:1)/1000))

par(mfcol = c(1, 2))
plot(S[,1], type="l")
plot(S[,2], type="l")
```

Create a mixing matrix A
```{r}
A <- matrix(c(0.5, 0.7, 0.423, 0.857), 2, 2)
```

Mix S using A
```{r}
X <- S %*% A
par(mfcol = c(1, 2))
plot(X[,1], type="l")
plot(X[,2], type="l")

```

Unmix using fastICA
```{r, include=FALSE}
a <- fastICA(X, 2, alg.typ = "parallel", fun = "logcosh", alpha = 1,
             method = "R", row.norm = FALSE, maxit = 200,
             tol = 0.0001, verbose = TRUE)
```


```{r}
par(mfcol = c(1, 2))
plot(1:500, a$S[,1], type = "l", xlab = "S'1", ylab = "")
plot(1:500, a$S[,2], type = "l", xlab = "S'2", ylab = "")
```


### ICA on the auto data
```{r, include=FALSE}
a <- fastICA(Auto_data_clust, 7, alg.typ = "parallel", fun = "logcosh", alpha = 1,
             method = "R", row.norm = FALSE, maxit = 200,
             tol = 0.0001, verbose = TRUE)
```

plot the independent components as a heatmap
```{r}
heatmap(a$S)
```


\newpage

## Homework

* I also did some alternative functions using different packages for my own reference and studying. Otherwise, the way covered in the lab and class is marked throughout the script.

```{r}
data(iris)
```

0. Subset the Iris dataset to only include `Sepal.Length`, `Sepal.Width`, `Petal.Length`, and `Petal.Width`. 

1. Write out the Kmeans algorithm by hand, and run two iterations of it. 

2. Run PCA on the Iris dataset. Plot a scatter plot of PC1 vs PC2 and include the percent variance those PCs describe. 

3. Run ICA on the Iris dataset. Plot the independent components as a heatmap.

4. Use Kmeans to cluster the Iris data. 
  * Use the silhouette function in the cluster package to find the optimal number of clusters for kmeans for the iris dataset. Then cluster using kmeans clustering. Does the data cluster by species? 
  * Using this clustering, color the PCA plot according to the clusters.
  
5. Use hierarchical clustering to cluster the Iris data.

  * Try two different linkage types, and two different distance metrics. 
  * For one linkage type and one distance metric, try two different cut points. 
  * Using this clustering, color the PCA plot according to the clusters. (6  plots in total)

____________________________________________________________________________________________________________________

0. subset iris data

```{r iris k-means}
data(iris)
as.data.frame(iris, row.names = NULL)

# dataset to cluster
iris_cluster <- iris[,1:4]
```

1. K-Means Algorithm by Hand

```{r k-means first iteration}
# pick 3 random points for centroids from data frame
set.seed(100)
centroid <- iris_cluster[sample(150,3),]
# empty matrix for squared distances to each centroid from each point to fill
dist_to_centroid <- matrix(NA,ncol=3,nrow=dim(iris_cluster)[1])

# k-means algorithm first iteration
centroid_new <- matrix(0,nrow=3,ncol=4)
# to identify where centroid is not replaced yet
while(all(centroid != centroid_new)){
# for all 150 rows in iris data and 3 centroids
  for(i in 1:150){
    for(j in 1:3){
      # calculating distance for each point to centroid
      dist_to_centroid[i,j] <- (sum((iris_cluster[i,] - centroid[j,])^2))
    }
  }
  # category is the clusters
  category <- rep(NA,150)
  for(i in 1:150){
    category[i] <- which.min(dist_to_centroid[i,])
  }
  # new centroid
  for(i in 1:3){
    centroid[i,] <- colMeans(iris_cluster[which(category==i),])
  }
  # update centroid values after first iteration
  centroid_new <- centroid
}
```


```{r k-means second iteration}
# k-means second iteration
# for all 150 rows in iris data and 3 centroids
  for(i in 1:150){
    for(j in 1:3){
      # calculating distance for each point to centroid
      dist_to_centroid[i,j] <- (sum((iris_cluster[i,] - centroid_new[j,])^2))
    }
  }
  category <- rep(NA,150)
  for(i in 1:150){
    category[i] <- which.min(dist_to_centroid[i,])
  }
  # new centroid
  for(i in 1:3){
    centroid_new[i,] <- colMeans(iris_cluster[which(category==i),])
  }
```

2. PCA (this is just another package I tried on my own; the way we did it in class is in the code chunk below)

```{r PCA}
# scale iris_cluster
iris_scale <- scale(iris_cluster)
#pca <- preProcess(x = iris_scale, method = 'pca', pcaComp = 2)
#iris_scale = predict(pca, iris_scale)

# PCA with percent variance of components
pca_iris <- princomp(iris_scale, cor = TRUE, scores = TRUE)
summary(pca_iris)

#plot(pca_iris, type = 'line')
# PC1 vs. PC2 plot
biplot(pca_iris)
```

Below is the PCA as demonstrated in class:

```{r PCA2}
iris_pca <- data.matrix(iris)

Center_iris <- apply(iris_pca, 2, function(x) x - mean(x))

Covariance_iris <- cov(Center_iris)

Eigen_value_iris <- eigen(Covariance_iris)$value

#columns are the eigen vectors
Eigen_vector_iris <- eigen(Covariance_iris)$vector

PC <- as.data.frame(data.matrix(Center_iris) %*% Eigen_vector_iris)

ggplot(PC, aes(PC[,1], PC[,2])) + geom_point(aes(PC[,1], PC[,2])) + labs(x = 'PC1 (92.48%)', y = 'PC2 (3.1%)')

round(cumsum(Eigen_value_iris)/sum(Eigen_value_iris) * 100, digits = 2)
```


3. ICA

```{r ICA}
S <- cbind(cos((1:500)/10), ((500:1)/1000))

par(mfcol = c(1, 2))
plot(S[,1], type="l")
plot(S[,2], type="l")

A <- matrix(c(0.5, 0.7, 0.423, 0.857), 2, 2)

X <- S %*% A
par(mfcol = c(1, 2))
plot(X[,1], type="l")
plot(X[,2], type="l")

a <- fastICA(iris_cluster, 4, alg.typ = "parallel", fun = "logcosh", alpha = 1,
             method = "R", row.norm = FALSE, maxit = 200,
             tol = 0.0001, verbose = TRUE)

heatmap(a$S)
```


4. K-Means on Iris Data

* The data does not cluster by species. The silhouette analysis produces optimal clusters of 2; there are 3 species.

```{r k-means clustering}
# silhouette analysis with cluster package
set.seed(20)
dis = dist(iris_cluster)^2
res = kmeans(iris_cluster,2)
sil = silhouette(res$cluster, dis)
plot(sil, col = res$cluster)
# pca plot colored by clusters
autoplot(prcomp(iris_cluster), data = iris_cluster, colour = res$cluster)

# alternative silhouette analysis
fviz_nbclust(iris_cluster, kmeans, method = "silhouette")

set.seed(29)
k2 <- kmeans(iris_cluster, 2, iter.max = 300)
clusplot(iris_cluster,
         k2$cluster,
         shade = TRUE,
         color = TRUE)
```

5. Hierarchical Clustering on Iris data

```{r hierarchical iris}
# euclidean distance and average linkage
hierarchical_iris <- dist(iris_cluster, method = "euclidean")
tree_iris <- hclust(hierarchical_iris, method="average")
plot(tree_iris)
# pca plot
cluster1 <- as.data.frame(cutree(tree_iris, k = 3))
iris <- cbind(iris, cluster1)
colnames(iris)[6] <- 'Cluster'
iris$Cluster <- as.factor(iris$Cluster)
autoplot(prcomp(iris_cluster), data = iris, colour = 'Cluster')

# cut tree at height 3
tree_iris2 <- hclust(hierarchical_iris, method="average")
plot(tree_iris2)
rect.hclust(tree_iris2, h = 3)
# pca plot
cluster2 <- as.data.frame(cutree(tree_iris2, k = 3))
iris <- cbind(iris, cluster2)
colnames(iris)[7] <- 'Cluster2'
iris$Cluster2 <- as.factor(iris$Cluster2)
autoplot(prcomp(iris_cluster), data = iris, colour = 'Cluster2')

# cut tree at height 1
tree_iris3 <- hclust(hierarchical_iris, method="average")
plot(tree_iris3)
rect.hclust(tree_iris3, h = 1)
# pca plot
cluster3 <- as.data.frame(cutree(tree_iris3, k = 3))
iris <- cbind(iris, cluster3)
colnames(iris)[8] <- 'Cluster3'
iris$Cluster3 <- as.factor(iris$Cluster3)
autoplot(prcomp(iris_cluster), data = iris, colour = 'Cluster3')

# manhattan distance and centroid linkage
hierarchical_iris2 <- dist(iris_cluster, method = "manhattan")
tree_iris4 <- hclust(hierarchical_iris2, method="centroid")
plot(tree_iris4)
rect.hclust(tree_iris4, k = 3)
# pca plot
cluster4 <- as.data.frame(cutree(tree_iris4, k = 3))
iris <- cbind(iris, cluster4)
colnames(iris)[9] <- 'Cluster4'
iris$Cluster4 <- as.factor(iris$Cluster4)
autoplot(prcomp(iris_cluster), data = iris, colour = 'Cluster4')

# euclidean distance and centroid linkage
tree_iris5 <- hclust(hierarchical_iris, method="centroid")
#tree_iris3 <- cutree(tree_iris3, k = 3)
plot(tree_iris5)
rect.hclust(tree_iris5, k = 3)
# pca plot
cluster5 <- as.data.frame(cutree(tree_iris5, k = 3))
iris <- cbind(iris, cluster5)
colnames(iris)[10] <- 'Cluster5'
iris$Cluster5 <- as.factor(iris$Cluster5)
autoplot(prcomp(iris_cluster), data = iris, colour = 'Cluster5')

# manhattan distance and average linkage
tree_iris6 <- hclust(hierarchical_iris2, method="average")
plot(tree_iris6)
rect.hclust(tree_iris6, k = 3)
# pca plot
cluster6 <- as.data.frame(cutree(tree_iris6, k = 3))
iris <- cbind(iris, cluster6)
colnames(iris)[11] <- 'Cluster6'
iris$Cluster6 <- as.factor(iris$Cluster6)
autoplot(prcomp(iris_cluster), data = iris, colour = 'Cluster6')

# manhattan distance average linkage with different cut
tree_iris7 <- hclust(hierarchical_iris2, method="average")
plot(tree_iris7)
rect.hclust(tree_iris7, h = 1)
# pca plot
cluster7 <- as.data.frame(cutree(tree_iris7, k = 3))
iris <- cbind(iris, cluster6)
colnames(iris)[12] <- 'Cluster7'
iris$Cluster6 <- as.factor(iris$Cluster7)
autoplot(prcomp(iris_cluster), data = iris, colour = 'Cluster7')

# alternative pca plot
res.pca <- PCA(iris_cluster, ncp = 2, graph = FALSE)
res.hcpc <- HCPC(res.pca, graph = FALSE)
fviz_cluster(res.hcpc,
             repel = TRUE,            # Avoid label overlapping
             show.clust.cent = TRUE, # Show cluster centers
             palette = "jco",         # Color palette see ?ggpubr::ggpar
             ggtheme = theme_minimal(),
             main = "Factor map"
             )
```


# Optional material
On PCA:

Eigen Vectors and Eigen Values http://www.visiondummy.com/2014/03/eigenvalues-eigenvectors/
Linear Algebra by Prof. Gilbert Strang https://ocw.mit.edu/courses/mathematics/18-06-linear-algebra-spring-2010/video-lectures/
http://www.cs.otago.ac.nz/cosc453/student_tutorials/principal_components.pdf
https://stats.stackexchange.com/questions/2691/making-sense-of-principal-component-analysis-eigenvectors-eigenvalues

On ICA: 

Independent Component Analysis: Algorithms and Applications https://www.cs.helsinki.fi/u/ahyvarin/papers/NN00new.pdf
Tutorial on ICA taken from http://rstudio-pubs-static.s3.amazonaws.com/93614_be30df613b2a4707b3e5a1a62f631d19.html



